---
title: "Project 8 Template"
output: pdf_document
---

```{r setup}
# Add to this package list for additional SL algorithms
pacman::p_load(
  tidyverse,
  ggthemes,
  ltmle,
  tmle,
  SuperLearner,
  tidymodels,
  caret,
  dagitty,
  ggdag,
  arm,
  xgboost,
  here,
  ggplot2,
  dplyr)

heart_disease <- read_csv("heart_disease_tmle.csv")

```

# Introduction

Heart disease is the leading cause of death in the United States, and treating it properly is an important public health goal. However, it is a complex disease with several different risk factors and potential treatments. Physicians typically recommend changes in diet, increased exercise, and/or medication to treat symptoms, but it is difficult to determine how effective any one of these factors is in treating the disease. In this project, you will explore SuperLearner, Targeted Maximum Likelihood Estimation (TMLE), and Longitudinal Targeted Maximum Likelihood Estimation (LTMLE). Using a simulated dataset, you will explore whether taking blood pressure medication reduces mortality risk. 

# Data

This dataset was simulated using R (so it does not come from a previous study or other data source). It contains several variables:

\begin{itemize}
    \item \textbf{blood\_pressure\_medication}: Treatment indicator for whether the individual took blood pressure medication (0 for control, 1 for treatment)
    \item \textbf{mortality}: Outcome indicator for whether the individual passed away from complications of heart disease (0 for no, 1 for yes)
    \item \textbf{age}: Age at time 1
    \item \textbf{sex\_at\_birth}: Sex assigned at birth (0 female, 1 male)
    \item \textbf{simplified\_race}: Simplified racial category. (1: White/Caucasian, 2: Black/African American, 3: Latinx, 4: Asian American, \newline 5: Mixed Race/Other)
    \item \textbf{income\_thousands}: Household income in thousands of dollars
    \item \textbf{college\_educ}: Indicator for college education (0 for no, 1 for yes)
    \item \textbf{bmi}: Body mass index (BMI)
    \item \textbf{chol}: Cholesterol level
    \item \textbf{blood\_pressure}: Systolic blood pressure 
    \item \textbf{bmi\_2}: BMI measured at time 2
    \item \textbf{chol\_2}: Cholesterol measured at time 2
    \item \textbf{blood\_pressure\_2}: BP measured at time 2
    \item \textbf{blood\_pressure\_medication\_2}: Whether the person took treatment at time period 2 
\end{itemize}

For the "SuperLearner" and "TMLE" portions, you can ignore any variable that ends in "\_2", we will reintroduce these for LTMLE.

# SuperLearner

## Modeling

Fit a SuperLearner model to estimate the probability of someone dying from complications of heart disease, conditional on treatment and the relevant covariates. Do the following:

\begin{enumerate}
    \item Choose a library of at least 5 machine learning algorithms to evaluate. \textbf{Note}: We did not cover how to hyperparameter tune constituent algorithms within SuperLearner in lab, but you are free to do so if you like (though not required to for this exercise). 
    \item Split your data into train and test sets.
    \item Train SuperLearner
    \item Report the risk and coefficient associated with each model, and the performance of the discrete winner and SuperLearner ensemble
    \item Create a confusion matrix and report your overall accuracy, recall, and precision
\end{enumerate}

```{r fit sl model}
# Fit SuperLearner Model

# Prepare data for SuperLearner.

## Drop the variables that have _2

heart_disease_sl <- heart_disease %>% 
  dplyr :: select(age, sex_at_birth, simplified_race, college_educ, income_thousands, bmi,
         blood_pressure, chol, blood_pressure_medication, mortality)
glimpse(heart_disease_sl)

## sl lib

listWrappers()

# Will use the following 5 algorithms: 1) SL.mean (to serve as a baseline), 2) SL.glmnet, 
# 3) SL.randomForest, 4) SL.xgboost, 5) SL.bayesglm
SL.library <- c("SL.mean", "SL.glmnet", "SL.randomForest", "SL.xgboost", "SL.bayesglm")

## Train/Test split

heart_disease_sl_split <-
  initial_split(heart_disease_sl, prop = 3/4) # create initial split

# Training 
# ----------
train <- # Declare the training set with rsample::training()
  training(heart_disease_sl_split)

# y_train 
y_train <- 
  train %>% 
  # pull and save as vector
  pull(mortality)    

# x_train  
x_train <-
  train %>%
  # drop the target variable 
  dplyr :: select(-mortality)   

# Testing 
# ----------
test <- 
  testing(heart_disease_sl_split)

# y test
y_test <- 
  test %>%
  pull(mortality)

# x test
x_test <- 
  test %>%
  dplyr :: select(-mortality) 

## Train SuperLearner

# set seed
set.seed(987)

# multiple models  
# ----------
sl = SuperLearner(Y = y_train,
                  X = x_train,
                  family = binomial(),
                  SL.library = SL.library)

## Risk and Coefficient of each model

sl$cvRisk
sl$coef

## Discrete winner and superlearner ensemble performance

discrete_winner <- which.min(sl$cvRisk)
winner_name <- SL.library[discrete_winner]

# predictions
# ----------
preds <- 
  predict(sl,             # use the superlearner not individual models
          x_test,         # prediction on test set
          onlySL = TRUE)  # use only models that were found to be useful (had weights)


# start with y_test
validation <- 
  y_test %>%
  # add our predictions - first column of predictions
  bind_cols(preds$pred[,1]) %>% 
  # rename columns
  rename(obs = `...1`,      # actual observations 
         pred = `...2`) %>% # predicted prob
  # change pred column so that obs above .5 are 1, otherewise 0
  mutate(pred = ifelse(pred >= .5, 
                           1,
                           0))

# view
head(validation)

## Confusion Matrix

caret::confusionMatrix(as.factor(validation$pred),
                       as.factor(validation$obs))

# Overall accuracy: 0.586
# Overall recall (aka sensistivity): 0.3598
# Overall precision (aka positive predictive value):  0.5955

```

## Discussion Questions

\begin{enumerate}
    \item Why should we, in general, prefer the SuperLearner ensemble to the discrete winner in cross-validation? Or in other words, what is the advantage of "blending" algorithms together and giving them each weights, rather than just using the single best algorithm (with best being defined as minimizing risk)?
\end{enumerate}

SuperLearner allows us to combine algorithms, which each pick up on different relationships. Doing this can help us reduce model bias and variance at the same time. On the other hand using the "best" single algorithm (defined by minimizing risk) could mean taht we choose an algorithm that had the lowest CV risk by chance, especially if risks are close. Choosing the single best model may also overfit to the validation set. SuperLearner reduces this by blending across models rather than fully committing to one.

# Targeted Maximum Likelihood Estimation

## Causal Diagram

TMLE requires estimating two models:

\begin{enumerate}
    \item The outcome model, or the relationship between the outcome and the treatment/predictors, $P(Y|(A,W)$.
    \item The propensity score model, or the relationship between assignment to treatment and predictors $P(A|W)$
\end{enumerate}

Using ggdag and daggity, draw a directed acylcic graph (DAG) that describes the relationships between the outcome, treatment, and covariates/predictors. Note, if you think there are covariates that are not related to other variables in the dataset, note this by either including them as freestanding nodes or by omitting them and noting omissions in your discussion.

Cindy note: I think all covariates are related to other variables in the dataset, which is reflected in the DAG below. I also use adjustmentSets() to help me identify variable sets that I can condition on to block non-causal paths without blocking the causal effect (given my covariates).

```{r dag1}
# DAG for TMLE
heart_dag <- dagitty('dag {
  mortality [outcome, pos="6,0"]
  blood_pressure_medication [exposure, pos="2,0"]
  age [pos="0,1"]
  sex_at_birth [pos="1,2"]
  simplified_race [pos="2,2"]
  income_thousands [pos="3,2"]
  college_educ [pos="4,2"]
  bmi [pos="1,1"]
  chol [pos="3,1"]
  blood_pressure [pos="4,1"]
  
  age -> bmi
  age -> blood_pressure
  age -> chol
  age -> mortality
  
  sex_at_birth -> bmi
  sex_at_birth -> blood_pressure
  sex_at_birth -> chol
  
  simplified_race -> blood_pressure
  simplified_race -> chol
  
  income_thousands -> college_educ
  income_thousands -> bmi
  
  college_educ -> bmi
  
  bmi -> blood_pressure
  bmi -> chol
  bmi -> mortality
  
  chol -> blood_pressure
  chol -> mortality
  
  blood_pressure -> blood_pressure_medication
  blood_pressure -> mortality
  
  blood_pressure_medication -> mortality
  
  income_thousands -> blood_pressure_medication
  college_educ -> blood_pressure_medication
}')

#Plot the DAG

heart_dag_tidy <- tidy_dagitty(heart_dag)
dag_plot <- ggdag(heart_dag_tidy, layout = "auto") + 
  theme_dag() +
  # Nodes with better visibility
  geom_dag_point(aes(color = name), size = 18, alpha = 0.7) +
  # Text styling: BLACK text only
  geom_dag_text(size = 3, 
                color = "black", 
                fontface = "bold",
                family = "sans") +  # Use sans-serif font for clarity
  # Improved edges
  geom_dag_edges(edge_width = 0.6, 
                arrow_directed = grid::arrow(length = grid::unit(7, "pt"), 
                                            type = "closed")) +
  scale_color_manual(values = c(
    "mortality" = "#FF5555",                
    "blood_pressure_medication" = "#5555FF", 
    "age" = "#55AA55",                      
    "sex_at_birth" = "#AA55AA",              
    "simplified_race" = "#CD8500",           
    "income_thousands" = "#00CDCD",         
    "college_educ" = "#FF69B4",             
    "bmi" = "#A52A2A",                     
    "chol" = "#FFD700",                     
    "blood_pressure" = "#8E8E8E"             
  )) +
  labs(title = "Directed Acyclic Graph (DAG) for Heart Disease Study",
       subtitle = "Relationships between mortality, blood pressure medication, and covariates",
       caption = "Note: Red = outcome, Blue = treatment, others = covariates") +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(face = "italic")
  )

# Display the plot
print(dag_plot)

# Identify adjustment sets for estimating the causal effect of treatment on outcome
adjustment_sets <- adjustmentSets(heart_dag, exposure = "blood_pressure_medication", outcome = "mortality", type = "minimal")
print("Minimal adjustment sets for estimating the effect of blood pressure medication on mortality:")
print(adjustment_sets)

```

## TMLE Estimation

Use the `tmle` package to estimate a model for the effect of blood pressure medication on the probability of mortality. Do the following:

\begin{enumerate}
    \item Use the same SuperLearner library you defined earlier
    \item Use the same outcome model and propensity score model that you specified in the DAG above. If in your DAG you concluded that it is not possible to make a causal inference from this dataset, specify a simpler model and note your assumptions for this step.
    \item Report the average treatment effect and any other relevant statistics
\end{enumerate}


AdjustemtnSets above allowed me to identify sets of variables that if I condition/adjust for, would block the non-causal paths without blocking the causal effect. Below I use TMLE and the following minimal adjustment set: age, blood_pressure, bmi, and chol. Note that this took a long time to run on my computer so I was unable to run the TMLE using the other two sets. However, I would have liked to compare the results from the three sets if I had enough time/computational power.
```{r tmle}

# Above, I identified three possible minimal adjustment sets:
# 1. { age, blood_pressure, bmi, chol }
# 2. { age, blood_pressure, bmi, sex_at_birth }
# 3. { blood_pressure, college_educ, income_thousands }
# I'll use the first set for this analysis 

# use heart_disease_sl
heart_disease_tmle <- heart_disease_sl

# Examine the structure
str(heart_disease_tmle)

# Make sure variables are properly formatted
heart_disease_tmle <- heart_disease_tmle %>%
  mutate(
    # Ensure binary variables are 0/1
    mortality = as.numeric(mortality),
    blood_pressure_medication = as.numeric(blood_pressure_medication),
    sex_at_birth = as.numeric(sex_at_birth),
    college_educ = as.numeric(college_educ),
    # Ensure simplified_race is a factor
    simplified_race = as.factor(simplified_race)
  )

# Define the SuperLearner library (same as above)
SL.library <- c("SL.mean", "SL.glmnet", "SL.randomForest", "SL.xgboost", "SL.bayesglm")

# Extract variables for TMLE
Y <- heart_disease_tmle$mortality  # Outcome
A <- heart_disease_tmle$blood_pressure_medication  # Treatment

# Extract covariates based on the first minimal adjustment set from our DAG
# { age, blood_pressure, bmi, chol }
W <- heart_disease_tmle %>%
  dplyr:: select(age, blood_pressure, bmi, chol)

# Double-check the data
summary(Y)
summary(A)
summary(W)

# Run TMLE
set.seed(42)  # For reproducibility

tmle_result <- tmle(
  Y = Y,  # Outcome: mortality
  A = A,  # Treatment: blood pressure medication
  W = W,  # Covariates: from minimal adjustment set
  Q.SL.library = SL.library,  # SuperLearner library for outcome model
  g.SL.library = SL.library,  # SuperLearner library for propensity score model
  family = "binomial"  # Binary outcome
)

# Print TMLE results
print(tmle_result)

# Extract the ATE
ate <- tmle_result$estimates$ATE
ate_ci <- c(tmle_result$estimates$ATE$CI[1], tmle_result$estimates$ATE$CI[2])
ate_pvalue <- tmle_result$estimates$ATE$pvalue

# Print the key results
cat("Average Treatment Effect (ATE):", round(ate$psi, 4), "\n")
cat("95% Confidence Interval:", round(ate_ci[1], 4), "to", round(ate_ci[2], 4), "\n")
cat("P-value:", round(ate_pvalue, 4), "\n")

# Interpretation based on the sign of the ATE
if (ate$psi < 0) {
  cat("The negative ATE suggests that blood pressure medication reduces mortality risk.\n")
} else if (ate$psi > 0) {
  cat("The positive ATE suggests that blood pressure medication increases mortality risk.\n")
} else {
  cat("The ATE is effectively zero, suggesting no effect of blood pressure medication on mortality.\n")
}



```

## Discussion Questions

\begin{enumerate}
    \item What is a "double robust" estimator? Why does it provide a guarantee of consistency if either the outcome model or propensity score model is correctly specified? Or in other words, why does mispecifying one of the models not break the analysis? \textbf{Hint}: When answering this question, think about how your introductory statistics courses emphasized using theory to determine the correct outcome model, and in this course how we explored the benefits of matching.
\end{enumerate}

# Double Robust Estimators

A double robust estimator combines an outcome regression model, which models the relationship between covariates, treatment, and outcome, with a propensity score model, which models the probability of treatment assignment given covariates. Double robust methods (like TMLE) use both models in a way that gives us "two chances" to get the right answer. When we use an outcome model approach, we need to correctly specify the outcome model to get consistent estimates. When we use a pure propensity score approach like inverse probability weighting, we need to correctly specify the propensity model. Double robust estimators combine these approaches mathematically so that if either the outcome model is correct or the propensity model is correct, the estimator is consistent. 

Traditional statistics courses emphasized theory-driven outcome modeling, focusing on getting the functional form right. Methods like matching instead focus on balancing covariates between treatment groups (like propensity score approaches). Double robust methods give us the best of both worlds because we can leverage our theoretical knowledge of outcome processes (as I did in the DAG step above) while also using propensity scores to account for selection bias. Instead of having to choose between these paradigms and being completely wrong if our choice fails, double robust methods allow us to be partially right and still get consistent estimates.

# LTMLE Estimation

Now imagine that everything you measured up until now was in "time period 1". Some people either choose not to or otherwise lack access to medication in that time period, but do start taking the medication in time period 2. Imagine we measure covariates like BMI, blood pressure, and cholesterol at that time for everyone in the study (indicated by a "_2" after the covariate name). 

## Causal Diagram

Update your causal diagram to incorporate this new information. \textbf{Note}: If your groups divides up sections and someone is working on LTMLE separately from TMLE then just draw a causal diagram even if it does not match the one you specified above.

\textbf{Hint}: Check out slide 27 from Maya's lecture, or slides 15-17 from Dave's second slide deck in week 8 on matching.

\textbf{Hint}: Keep in mind that any of the variables that end in "\_2" are likely affected by both the previous covariates and the first treatment when drawing your DAG.

```{r dag2}
# DAG for LTMLE

# Create a simplified longitudinal DAG with two time periods
long_dag <- dagitty('dag {
  mortality [outcome, pos="6,0"]
  blood_pressure_medication [exposure, pos="3,0"]
  age [pos="0,2"]
  sex_at_birth [pos="0,1"]
  simplified_race [pos="0,0"]
  income_thousands [pos="0,-1"]
  college_educ [pos="0,-2"]
  bmi [pos="1.5,1"]
  chol [pos="1.5,0"]
  blood_pressure [pos="1.5,-1"]
  
  blood_pressure_medication_2 [exposure, pos="4.5,0"]
  bmi_2 [pos="3,1"]
  chol_2 [pos="3,0"]
  blood_pressure_2 [pos="3,-1"]
  
  age -> bmi
  age -> blood_pressure
  age -> chol
  sex_at_birth -> bmi
  sex_at_birth -> blood_pressure
  sex_at_birth -> chol
  simplified_race -> blood_pressure
  simplified_race -> chol
  income_thousands -> college_educ
  income_thousands -> bmi
  college_educ -> bmi
  
  bmi -> blood_pressure
  bmi -> chol
  blood_pressure -> blood_pressure_medication
  chol -> blood_pressure_medication
  
  income_thousands -> blood_pressure_medication
  college_educ -> blood_pressure_medication
  
  blood_pressure_medication -> bmi_2
  blood_pressure_medication -> chol_2
  blood_pressure_medication -> blood_pressure_2
  
  bmi -> bmi_2
  chol -> chol_2
  blood_pressure -> blood_pressure_2
  
  blood_pressure_medication -> blood_pressure_medication_2
  
  bmi_2 -> blood_pressure_medication_2
  chol_2 -> blood_pressure_medication_2
  blood_pressure_2 -> blood_pressure_medication_2
  
  age -> mortality
  bmi -> mortality
  chol -> mortality
  blood_pressure -> mortality
  blood_pressure_medication -> mortality
  
  bmi_2 -> mortality
  chol_2 -> mortality
  blood_pressure_2 -> mortality
  blood_pressure_medication_2 -> mortality
}')

# Prepare for plotting with ggdag
long_dag_tidy <- tidy_dagitty(long_dag)

# Set up the color scheme for different node types
time1_vars <- c("age", "sex_at_birth", "simplified_race", "income_thousands", 
                "college_educ", "bmi", "chol", "blood_pressure", "blood_pressure_medication")
time2_vars <- c("bmi_2", "chol_2", "blood_pressure_2", "blood_pressure_medication_2")
outcome_var <- "mortality"

# Create the ggdag plot with clear time periods
ggdag_plot <- ggdag(long_dag_tidy) +
  # Use different node shapes for different time periods
  geom_dag_point(aes(color = name, 
                     shape = ifelse(name %in% time2_vars, "Time 2", 
                                    ifelse(name == outcome_var, "Outcome", "Time 1"))),
                 size = 14, alpha = 0.7) +
  
  # Custom node shapes
  scale_shape_manual(name = "Time Period",
                     values = c("Time 1" = 16, "Time 2" = 17, "Outcome" = 18)) +
  
  # Use clear black text for all nodes
  geom_dag_text(size = 3, color = "black", fontface = "bold") +
  
  # Customize arrows
  geom_dag_edges(edge_width = 0.5, 
                 arrow_directed = grid::arrow(length = grid::unit(6, "pt"), type = "closed")) +
  
  # Set node colors
  scale_color_manual(values = c(
    "mortality" = "#FF5555",                
    "blood_pressure_medication" = "#5555FF", 
    "blood_pressure_medication_2" = "#0000FF",
    "age" = "#55AA55",                      
    "sex_at_birth" = "#AA55AA",              
    "simplified_race" = "#CD8500",           
    "income_thousands" = "#00CDCD",         
    "college_educ" = "#FF69B4",             
    "bmi" = "#A52A2A",                     
    "bmi_2" = "#8B0000",
    "chol" = "#FFD700",                     
    "chol_2" = "#DAA520",
    "blood_pressure" = "#8E8E8E",
    "blood_pressure_2" = "#4D4D4D"
  )) +
  
  # Add a clear title and subtitle
  labs(title = "Longitudinal DAG for Heart Disease Study",
       subtitle = "Time-varying treatments, covariates, and mortality",
       caption = "Circle = Time 1, Triangle = Time 2, Diamond = Outcome") +
  
  # Theme adjustments
  theme_dag() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(face = "italic", hjust = 0)
  )

# Print the plot
print(ggdag_plot)

# Create an even more focused version with time period boundaries
# First create a more structured layout with clearer time periods
structured_dag <- dagitty('dag {
  mortality [outcome, pos="5,0"]
  blood_pressure_medication [exposure, pos="2,0"]
  age [pos="0,2"]
  sex_at_birth [pos="0,1"]
  bmi [pos="1,2"]
  chol [pos="1,1"]
  blood_pressure [pos="1,0"]
  
  blood_pressure_medication_2 [exposure, pos="4,0"]
  bmi_2 [pos="3,2"]
  chol_2 [pos="3,1"]
  blood_pressure_2 [pos="3,0"]
  
  age -> bmi
  age -> blood_pressure
  age -> mortality
  sex_at_birth -> bmi
  
  bmi -> blood_pressure
  bmi -> blood_pressure_medication
  blood_pressure -> blood_pressure_medication
  
  blood_pressure_medication -> bmi_2
  blood_pressure_medication -> blood_pressure_2
  blood_pressure_medication -> chol_2
  blood_pressure_medication -> blood_pressure_medication_2
  
  bmi -> bmi_2
  chol -> chol_2
  blood_pressure -> blood_pressure_2
  
  bmi_2 -> blood_pressure_2
  bmi_2 -> blood_pressure_medication_2
  chol_2 -> blood_pressure_medication_2
  blood_pressure_2 -> blood_pressure_medication_2
  
  bmi -> mortality
  blood_pressure -> mortality
  blood_pressure_medication -> mortality
  bmi_2 -> mortality
  blood_pressure_2 -> mortality
  blood_pressure_medication_2 -> mortality
}')

# Convert to tidy format
structured_tidy <- tidy_dagitty(structured_dag)

# Prepare the time period overlay
time1_box <- data.frame(
  x = c(-0.5, 2.5, 2.5, -0.5, -0.5),
  y = c(-0.5, -0.5, 2.5, 2.5, -0.5)
)

time2_box <- data.frame(
  x = c(2.5, 4.5, 4.5, 2.5, 2.5),
  y = c(-0.5, -0.5, 2.5, 2.5, -0.5)
)

# Create the final plot with time period boundaries
final_dag_plot <- ggplot() +
  # Add time period backgrounds
  geom_polygon(data = time1_box, aes(x = x, y = y), fill = "lightblue", alpha = 0.2) +
  geom_polygon(data = time2_box, aes(x = x, y = y), fill = "lightgreen", alpha = 0.2) +
  
  # Add time period labels
  annotate("text", x = 1, y = 2.3, label = "Time Period 1", fontface = "bold", size = 4) +
  annotate("text", x = 3.5, y = 2.3, label = "Time Period 2", fontface = "bold", size = 4) +
  
  # Add DAG edges
  geom_dag_edges(data = structured_tidy,
                 aes(x = x, y = y, xend = xend, yend = yend),
                 edge_width = 0.5,
                 arrow = grid::arrow(length = grid::unit(6, "pt"), type = "closed")) +
  
  # Add nodes with different shapes by time period
  geom_dag_point(data = subset(structured_tidy, name != "edges" & 
                              !grepl("_2$", name) & 
                              name != "mortality"),
                aes(x = x, y = y, fill = name),
                size = 15, alpha = 0.7, shape = 21) +  # Circle for Time 1
  
  geom_dag_point(data = subset(structured_tidy, name != "edges" & 
                              grepl("_2$", name)),
                aes(x = x, y = y, fill = name),
                size = 15, alpha = 0.7, shape = 22) +  # Square for Time 2
  
  geom_dag_point(data = subset(structured_tidy, name == "mortality"),
                aes(x = x, y = y, fill = name),
                size = 15, alpha = 0.7, shape = 23) +  # Diamond for outcome
  
  # Add clear black text
  geom_dag_text(data = structured_tidy,
                aes(x = x, y = y, label = name),
                size = 3, color = "black", fontface = "bold") +
  
  # Set node colors
  scale_fill_manual(values = c(
    "mortality" = "#FF5555",                
    "blood_pressure_medication" = "#5555FF", 
    "blood_pressure_medication_2" = "#0000FF",
    "age" = "#55AA55",                      
    "sex_at_birth" = "#AA55AA",              
    "bmi" = "#A52A2A",                     
    "bmi_2" = "#8B0000",
    "chol" = "#FFD700",                     
    "chol_2" = "#DAA520",
    "blood_pressure" = "#8E8E8E",
    "blood_pressure_2" = "#4D4D4D"
  )) +
  
  # Apply DAG theme and customizations
  theme_dag() +
  
  # Add titles
  labs(title = "Longitudinal DAG for Heart Disease Study",
       subtitle = "Time-varying treatments, covariates, and mortality",
       caption = "Circle = Time 1, Square = Time 2, Diamond = Outcome
                 Blue rectangles = Time Period 1, Green rectangles = Time Period 2") +
  
  # Theme adjustments
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(face = "italic", hjust = 0)
  )

# Print the final plot
print(final_dag_plot)

# Identify adjustment sets for time-varying treatment effects
# For LTMLE, we need to consider the sequential adjustment sets
adjustment_time1 <- adjustmentSets(
  structured_dag,
  exposure = "blood_pressure_medication",
  outcome = c("mortality", "bmi_2", "chol_2", "blood_pressure_2", "blood_pressure_medication_2")
)

adjustment_time2 <- adjustmentSets(
  structured_dag,
  exposure = "blood_pressure_medication_2",
  outcome = "mortality"
)

print("Adjustment sets for time 1 treatment effect:")
print(adjustment_time1)

print("Adjustment sets for time 2 treatment effect:")
print(adjustment_time2)
```

## LTMLE Estimation

Use the `ltmle` package for this section. First fit a "naive model" that \textbf{does not} control for the time-dependent confounding. Then run a LTMLE model that does control for any time dependent confounding. Follow the same steps as in the TMLE section. Do you see a difference between the two estimates?

```{r ltmle}
# naive TMLE (no time-dependent confounding control)

# Joint treatment (e.g., ever treated over two waves)
heart_disease$ever_treated <- ifelse(
  heart_disease$blood_pressure_medication == 1 | heart_disease$blood_pressure_medication_2 == 1, 1, 0
)

# Naive TMLE model
naive_tmle <- tmle(
  Y = heart_disease$mortality,
  A = heart_disease$ever_treated,
  W = heart_disease[, c("age", "sex_at_birth", "simplified_race", "college_educ", "income_thousands")],
  family = "binomial",
  Q.SL.library = SL.library,
  g.SL.library = SL.library
)
summary(naive_tmle)


#LTMLE

library(ltmle)

# Prepare dataset in wide format
ltmle_data <- heart_disease[, c(
  "age", "sex_at_birth", "simplified_race", "college_educ", "income_thousands",  # Baseline covariates
  "bmi", "blood_pressure", "chol",                   # Time 1 covariates
  "blood_pressure_medication",                      # Time 1 treatment
  "bmi_2", "blood_pressure_2", "chol_2",            # Time 2 covariates
  "blood_pressure_medication_2",                    # Time 2 treatment
  "mortality"                                       # Outcome
)]

# Specify node types
Anodes <- c("blood_pressure_medication", "blood_pressure_medication_2")
Lnodes <- c("bmi", "blood_pressure", "chol", "bmi_2", "blood_pressure_2", "chol_2")
Ynodes <- "mortality"

abar <- list(
  control = c(0, 0),   # Never treated at either time
  treated = c(1, 1)    # Treated at both time points
)


# Fit LTMLE model
ltmle_fit <- ltmle(
  data = ltmle_data,
  Anodes = Anodes,
  Lnodes = Lnodes,
  Ynodes = Ynodes,
  survivalOutcome = FALSE,
  SL.library = c("SL.mean", "SL.glm", "SL.bayesglm"),
  abar = abar
)
summary(ltmle_fit)


```

Above I first tried to use my fully defined SL library for the LTMLE model as I defined the library for my naive model and TMLE above, but after leaving the code running for 7 hours, I stopped it, removed the more computationally intensive/slow learners: SL.glmnet, SL.randomForest, and SL.xgboost, and I just included "SL.mean", "SL.glm", "SL.bayesglm". After removing these learners, I see that LTMLE model shows a positive treatment effect (0.38) while the naive model shows a negative treatment effect (Additive Effect Parameter Estimate: -0.18309). The fact that adjusting for time-dependent confounding reversed the effect suggests that there was substantial time-dependent confounding. There was also a big change in the magnitude of the effect (odds ratio incerases from 0.48, harmful, to 5.67, beneficial).


## Discussion Questions

\begin{enumerate}
    \item What sorts of time-dependent confounding should we be especially worried about? For instance, would we be concerned about a running variable for age the same way we might be concerned about blood pressure measured at two different times?
\end{enumerate}

I'd be especially worried about health measures or biomarkers that both respond to treatment and influence subsequent treatment decisions, because these create feedback loops that bias standard analyses. Blood pressure is a good example of this because it's affected by medication at time 1, then influences medication decisions at time 2, while also independently affecting mortality risk. This creates complex pathways that standard methods cannot get at (conventional approaches might introduce collider bias by adjusting for variables on the causal pathway). A running variable like age progresses independently of treatment. While age affects treatment decisions and outcomes, treatment doesn't affect aging itself, making it different from true time-varying confounders like blood pressure, BMI, or cholesterol. I think the most problematic time-dependent confounders are those that serve as both outcomes of previous treatments and determinants of future treatments, particularly when they're strong predictors of the final outcome. 

